### 2020-11-21 pldi21/case-studies.tex
finished tweaking sec 2 and expanded filter case study substantially

Next, the instructor wants to ``eyeball'' reasonable \li{cutoffs} between letter grades 
by directly manipulating a highly domain-specific livelit, \li{\$grade_cutoffs}, that provides draggable ``paddles'' 
superimposed on a live visualization of the distribution of the \li{averages} value provided as a parameter 
(for the sake of space, we plot dots rather than a histogram).
% The value of \li{cutoffs} is a labeled 4-tuple containing each cut-off.

@@

\subsubsection{Expansion Typing} 
\label{sec:expansion-typing}
To support abstract reasoning about the type of the expansion,
livelit definitions declare an \emph{expansion type}.
The declarations of the two livelits in Fig.~\ref{fig:grading},
eliding their implementations, are:
\begin{lstlisting}[numbers=none,xleftmargin=0cm]
livelit $dataframe at Dataframe {...}
livelit $grade_cutoffs(averages: List(Float)) at 
  (.A Float, .B Float, .C Float, .D Float) {...}
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe},
which classifies tabular floating point data together with string row and column names (see the expansion above).
The expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (field labels are written \li{.label}
rather than \li{label:} in Hazel).
Hazel displays the information in the livelit declaration when the cursor is on the livelit's name,
just as it displays typing information in other situations (not shown).\todo{cite HATRA}{}

@@

Livelit abbreviations can partially apply parameters. For example, consider the slider livelit
from Fig.~\ref{fig:color}\todo{where?}{}\todo{int or float?}:
\begin{lstlisting}[numbers=none,xleftmargin=0cm]
livelit $slider (min: Int) (max: Int) at Int {...}
\end{lstlisting}
We can partially apply the first parameter to define a parameterized non-negative slider livelit:
\begin{lstlisting}[numbers=none,xleftmargin=0cm]
let $nnslider = $slider 0 in ...
\end{lstlisting}
Only livelits with no remaining parameters can be invoked, 
so writing \li{\$nnslider} in expression position will display as a ``missing livelit parameter'' error.%
\footnote{\label{footnote:typing}In Hazel, erroneous expressions 
are automatically placed inside holes so that they do not prevent other parts of the program from evaluating
\cite{HazelnutLive}.}

@@

\paragraph{Capture Avoidance}
Splices and parameters can appear anywhere in the expansion. 
This becomes potentially problematic when
the expansion places the parameter or splice under a binder, e.g. in the body of a function or \li{let} binding.
Na\"ively, this could cause inadvertent capture of the bound variable by a free variable
in the parameter or splice. For example, consider a livelit that generates an expansion
of the following seemingly innocuous form:
\begin{lstlisting}[numbers=none]
let len = strlen <splice1> in
Some (<splice2> + len)
\end{lstlisting}
Here, \li{<splice2>} appears under the binding of \li{len}. If the client has filled
\li{<splice2>} with an expression that refers to a client-side binding of \li{len},
these references would na\"ively be captured. This would not occur in \li{<splice1>}, 
because the \li{let} is not recursive.
This breaks abstraction and is notoriously difficult to debug,
both for the livelit provider, who has no way to predict which variables a client will use,
 and the client, who does not know which variables the provider used.

@@

\paragraph{Context Independence}
The example expansion above used a library function, \li{strlen}.
Na\"ively, this expansion would break if placed 
in client contexts where \li{strlen} is not bound, or bound to 
an unexpected value.
To avoid requiring clients to determine and satisfy these invisible 
dependencies, the livelits mechanism enforces \emph{context independence}:
generated expansions are valid in any context. Dependencies are bound 
relative to the livelit definition site (see Sec.~\ref{sec:expansion}).

@@

\subsubsection{Closure Collection} The subtlety here is that 
evaluation in Hazel is defined for closed expressions as is typical,
but parameters and splices can be open, i.e. refer to variables in the surrounding
context. In order to provide a suitable run-time environment that binds these variables, 
Hazel performs \textbf{closure collection} in two phases.

In the first phase, \emph{proto-closure collection}, 
Hazel replaces each livelit with a uniquely numbered hole and then evaluates the program 
using the semantics for evaluating programs with holes developed by \citet{HazelnutLive}.
Evaluation proceeds around these holes, producing a result with hole closures, i.e. holes with environments.

For example, there is one closure for \li{\$dataframe} in Fig.~\ref{fig:grading}.
It contains the value of the \li{q1max}\todo{what?}{} variable from Line 1 and any other variables in the context
around that livelit. 
These values can be used to evaluate 
splices that use the corresponding variables, such as the cell selected in Fig.~\ref{fig:grading}(c).

Similarly, the closure for \li{\$grade_cutoffs} in Fig.~\ref{fig:grading} includes a result for
the \li{grades} variable. However, there is a problem: this variable's result depends on the
expansion of the \li{\$dataframe} livelit. If we stop after the first phase of closure collection,
where livelits have not yet been expanded,
then no useful value for the \li{grades} variable will be available after proto-closure collection:
it will just be a \li{\$dataframe} livelit hole closure.
For this reason, there is a second phase of closure collection, called \emph{closure resumption}, 
where any livelit holes that appear
in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated
(in this case, for the \li{\$dataframe} livelit), the hole is filled, 
and evaluation resumes.
Livelit expansions do not contain livelits, so no subsequent resumption phases are necessary.

Hazel does not need to evaluate the program multiple times in order to support live closure collection
as well as the usual live evaluation services it offers, because the final evaluation result 
can also be determined by resumption.
(We discuss how to address subtleties that arise 
in the presence of non-commutative side effects in Sec.~\ref{sec:calculus-closure-collection}.)

\subsubsection{Indeterminate Results}
We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully:
splices and parameters can be incomplete, i.e. they can contain holes, so
evaluation does not always result in a value \cite{HazelnutLive}.
In particular, when a hole appears in elimination position, e.g. in function position of a function application,
 the result is instead an indeterminate expression.
When a livelit requests an evaluation result, it must be able to handle indeterminate results.
For example, if the parameter to \li{\$grade_cutoffs} were a hole or some other indeterminate expression,
then it would have degraded functionality
(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements).
We will return to how livelit implementations handle this situation in Sec.~\ref{sec:live-evaluation-def}.

\subsubsection{Case Study: Live Image Transformations}\label{sec:image-transformation}
\todo{revise this once we have final figure}{}
Our next case study is a simple image transformation livelit, \li{\$img_filter},
demonstrated in Fig.~\ref{fig:img-transformation}. This livelit takes 
a URL to an image as a parameter, and contains two splices of type \li{Int},
one to adjust the contrast, and the other to adjust the brightness.
In this example, we have filled those splices with a \li{\$slider}, but 
as above we could enter any expression of type \li{Int}.
The livelit shows a live preview of the transformed image.
The expansion generates the necessary calls to image processing functions, 
not shown.

The novelty in this example is that the livelit appears inside a function, 
\li{classic}, which takes an image URL as input. At the bottom 
of the figure, \li{classic} is mapped over a list (here, containing only 
two elements for the sake of presentation). Consequently, there are now 
two closures associated with the application of \li{\$img_filter} in the 
body of \li{classic}. Rather than disabling live evaluation in situations like 
this, Hazel instead allows the programmer to select between the closures when 
the cursor is on the livelit expression (via a simple sidebar toggle, not shown, which is shared 
with the general closure selection UI \cite{HazelnutLive}). 
The live feedback is based on the selected closure.
This makes it easy to see how the image filter being designed here will affect a
number of example images by quickly toggling between closures. The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.

>>>>

Next, the instructor wants to ``eyeball'' reasonable \li{cutoffs} between letter grades 
by directly manipulating a domain-specific livelit, \li{\$grade_cutoffs}, that provides draggable ``paddles'' 
superimposed on a live visualization of the distribution of \li{averages}, which is provided as a livelit parameter.
% The value of \li{cutoffs} is a labeled 4-tuple containing each cut-off.

@@

\subsubsection{Expansion Typing} 
\label{sec:expansion-typing}
To support abstract reasoning about the type of the expansion,
livelit definitions declare an \emph{expansion type}.
The declarations of the livelits in Fig.~\ref{fig:grading},
eliding their implementations, are:
\begin{lstlisting}[numbers=none,xleftmargin=0cm]
livelit $dataframe at Dataframe {...}
livelit $grade_cutoffs(averages: List(Float)) at 
  (.A Float, .B Float, .C Float, .D Float) {...}
livelit $slider (min: Int) (max: Int) at Int {...}
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe},
which classifies tabular floating point data together with string row and column names (see the expansion above).
The expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (field labels are written \li{.label}
rather than \li{label:} in Hazel).
Hazel displays the information in the livelit declaration when the cursor is on the livelit's name,
just as it displays typing information in other situations (not shown).\todo{cite HATRA}{}

@@

Livelit abbreviations can partially apply parameters. For example, 
we can partially apply the first parameter of \li{\$slider} to define a parameterized unsigned slider livelit:
\begin{lstlisting}[numbers=none,xleftmargin=0cm]
let $uslider = $slider 0 in ...
\end{lstlisting}
Only livelits with no remaining parameters can be invoked, 
so writing \li{\$uslider} in expression position will display as a ``missing livelit parameter'' error.%
\footnote{\label{footnote:typing}In Hazel, erroneous expressions 
are automatically placed inside holes so that they do not prevent other parts of the program from evaluating
\cite{HazelnutLive}.}

@@

\textbf{Capture Avoidance.}
Splices and parameters can appear anywhere in the expansion. 
This becomes problematic when
they appear under a binder, e.g. in the body of a function or \li{let} binding.
Na\"ively, this could cause inadvertent capture of the bound variable by a free variable
in the parameter or splice. For example, consider a livelit that generates an expansion
of the following seemingly innocuous form:
\begin{lstlisting}[numbers=none]
let len = strlen <splice1> in
Some (<splice2> + len)
\end{lstlisting}
Here, \li{<splice2>} appears under the binding of \li{len}. If the client has filled
\li{<splice2>} with an expression that refers to a client-side binding of \li{len},
these references would na\"ively be captured. This would not occur in \li{<splice1>}, 
because the \li{let} is not recursive.
This breaks abstraction and is notoriously difficult to debug,
both for the livelit provider, who has no way to predict which variables a client will use,
 and the client, who does not know which variables the provider used.

@@

\textbf{Context Independence.}
The example expansion above used a library function, \li{strlen}.
Na\"ively, this expansion would break if placed 
in client contexts where \li{strlen} is not bound, or bound to 
an unexpected value.
To avoid requiring clients to determine and satisfy these invisible 
dependencies, the livelits mechanism enforces \emph{context independence}:
generated expansions are valid in any context. Dependencies are bound 
relative to the livelit definition site (see Sec.~\ref{sec:expansion}).

@@

\subsubsection{Closure Collection} The subtlety is that 
evaluation in Hazel is defined for closed expressions as usual,
but parameters and splices can be open, i.e. refer to surrounding variables. 
To provide a environment that binds these variables, 
Hazel performs \textbf{closure collection} in two phases.

In the first phase, \emph{proto-closure collection}, 
Hazel replaces each livelit with a uniquely numbered hole and then evaluates the program 
using the semantics for evaluating programs with holes developed by \citet{HazelnutLive}.
Evaluation proceeds around these holes, producing a result containing 
corresponding hole closures, i.e. holes with environments.

For example, there is one closure for \li{\$dataframe} in Fig.~\ref{fig:grading}.
It contains the value of \li{q1max}\todo{what?}{} and the other variables in scope. 
These values can be used to evaluate 
splices that use the corresponding variables, such as the cell selected in Fig.~\ref{fig:grading}(c).

Similarly, the closure for \li{\$grade_cutoffs} in Fig.~\ref{fig:grading} includes 
the necessary \li{averages} variable, but 
its value depends on \li{grades}, which is determined by \li{\$dataframe}. 
If we stop after proto-closure collection,
then no useful value will be available:
the result will be \emph{indeterminate}, because \li{\$dataframe} has been replaced with a hole \cite{HazelnutLive}.
For this reason, there is a second phase of closure collection, called \emph{closure resumption}, 
where any livelit holes that appear
in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated
(in this case, for the \li{\$dataframe} livelit) and evaluation resumes.
% Livelit expansions do not contain livelits, so no subsequent expansion / resumption phases are necessary.

% Hazel does not need to evaluate the program multiple times in order to support live closure collection
% as well as the usual live evaluation services it offers, because the final evaluation result 
% can also be determined by resumption.
% (We discuss how to address subtleties that arise 
% in the presence of non-commutative side effects in Sec.~\ref{sec:calculus-closure-collection}.)

\subsubsection{Indeterminate Results}
Even after resumption, some elements of the closure may remain indeterminate, e.g. due to holes that 
are not filled with livelits. 
When a livelit requests an evaluation result, it must be able to handle these indeterminate results.
For example, if there were missing grades,
then \li{\$grade_cutoffs} would have degraded functionality: 
it would display only the list elements that are values on the timeline, skipping indeterminate elements.
We will return to how this occurs in Sec.~\ref{sec:live-evaluation-def}.

\subsubsection{Case Study: Live Photo Filters}\label{sec:image-transformation}
\todo{revise this once we have final figure}{}
Closure collection can produce multiple closures when, for example, 
a livelit appears inside a function applied multiple times. 
Our next case study, in the photography domain,
demonstrates how Hazel handle this situation, and 
the workflow it enables.

We interviewed a photographer 
who described a typical workflow: 
they use the Lightroom application to 
apply a set of adjustments and filters 
across all photos in a collection before making 
individual adjustments. 
Many photographers do this one photo at a time,
though this photographer had recently learned how to
use Lightroom's saved presets to 
apply adjustments to multiple photos at once.
However, they remained dissatisfied by the workflow.
They wanted to be able to see how the shared settings affected 
multiple photos as they tweaked them, without having to 
save and apply the preset repeatedly.
They also wanted to be able to change the shared settings
even after making certain individual adjustments.
They also expressed interest in defining parameterized 
presets, and in automating the process of 
posting sets to a website and social media.

% For our next case study, we 
% developed a library of 
% image transformation livelits, 
% motivated by a desire to automate

% The livelits we have considered so far generate simple values.

Motivated by this interview, 
we prototyped a collection of image transformation livelits.
One of these, \li{\$something} is demonstrated in Fig.~\ref{fig:img-transformation},
within a function, \li{classic_look}, that serveas as a ``preset'' that is 
mapped over a list of images (loaded by URL) at the bottom of the figure. 
% We show two for space reasons.

...\todo{finalize?}{}
This livelit takes 
a URL to an image as a parameter, and contains two splices of type \li{Int},
one to adjust the contrast, and the other to adjust the brightness.
In this example, we have filled those splices with a \li{\$slider}, but 
as above we could enter any expression of type \li{Int}, e.g. 
an arithmetic expression or a variable. 
The livelit shows a live preview of the transformed image.
The expansion generates the necessary calls to an image processing library, 
not shown.

Because the livelit appears inside a function applied (by \li{map}) twice, 
there are now two closures associated with the livelit. 
Rather than disabling live evaluation in situations like 
this, Hazel instead allows the programmer to select between the closures when 
the cursor is on the livelit expression via a simple sidebar toggle. 
This causes the view to change as shown in ...\todo{todo}{}
The live feedback is based on the selected closure.
This makes it easy to see how the image filter being designed here will affect a
number of example images by quickly toggling between closures. 
The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.

We showed this and similar examples to the photographer we had interviewed. 
They expressed substantial interest in this approach despite having only 
limited programming experience (with Python). They made the fair point that it 
would take substantial effort to match Lightroom's breadth of filters and effects, 
but with sustained effort, they agreed that this approach could be more powerful than 
Lightroom's point-and-click interface while retaining many of its benefits (specifically
mentioning the graphical sliders). Although this was only a single interview,
it is consistent with the body of evidence summarized in Sec.~\ref{sec:background}.
